Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-04 paper/paper.tex
update codes and RJournal layout, not finishing...

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here
\usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs} \usepackage[ruled,vlined, linesnumbered]{algorithm2e} \usepackage{todonotes}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Visual diagnostics for constrained optimisation with application to guided tours}

  \author{
        H.Sherry Zhang \thanks{The authors gratefully acknowledge \ldots{}} \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Author 2 \\
    Department of ZZZ, University of WWW\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Visual diagnostics for constrained optimisation with application to guided tours}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Projection pursuit searches for interesting low-dimensional views of high-dimensional data via the optimisation of an index function. The initial paper by Friedman \& Tukey in 1974 stated that ``the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure.'' However, while many projection pursuit indices have been proposed in the literature, less work has been done on the optimisation procedures. In this paper we introduce visual diagnostics for optimisation algorithms, in particular those available in the projection pursuit guided tour. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, ferrn, has been created to implement the diagnostics.
\end{abstract}

\noindent%
{\it Keywords:} optimisation, projection pursuit, guided tour, visualisation, diagnostics, R
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

Mathematical details of the geodesic interpolation can be found in \citet{buja2005computational}. Figure \ref{fig:tour-path} shows a sketch of the tour path. The blue frames are produced by the projection pursuit optimisation algorithm, and the white frames interpolate between them. The tour method has been implemented in the \texttt{tourr} package in R, available on the Comprehensive R Archive Network at \url{https://cran.r-project.org/web/packages/tourr/} \citep{wickham2011tourrpackage}.

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth,height=0.3\textheight]{/Users/pmen0008/Desktop/paper-tour-vis/img/tour_path}

@@

\caption{Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.}\label{fig:tour-path}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-search-1}

@@

\caption{A comparison of \texttt{search\_better} with different values of the parameter \texttt{max\_tries}. A six-variable dataset \texttt{boa6} is used with \texttt{holes} index on a 2D problem. \texttt{max\_tries\ =\ 500} allows the optimiser to find better basis with higher index value at iteration six.}\label{fig:toy-search}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-interp-1}

@@

\caption{The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies.}\label{fig:toy-interp}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-pca-1}

@@

\caption{1D projection on the 5-variable dataset \texttt{boa5} with two optimisers: \texttt{search\_better} and \texttt{search\_geodesic}. The yellow point corresponds to the theoretical best basis {[}0, 1, 0, 0, 0{]} with \texttt{V2} being the only non-normal variable in the dataset. The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \texttt{search\_geodesic} has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.}\label{fig:toy-pca}
\end{figure}

@@

Figure \ref{fig:toy-pca} plots the first two principal components of two search paths, one using \texttt{search\_better} and the other using \texttt{search\_geodesic}. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in \texttt{search\_better} is broader, controlled by \texttt{alpha} parameter, which is default to 0.5 while the directional search in \texttt{search\_geodesic}, controlled by \texttt{delta} with a default of 0.01, is so tiny that it can barely be seen.

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-pca-animated-1}

@@

\caption{A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm.}\label{fig:toy-pca-animated}
\end{figure}

Animated plots can be informative in diagnostics, especially in the case of PCA plot when the starting and ending of the search is not clear. Figure \ref{fig:toy-pca-animated} shows six frames of an animated version of Figure \ref{fig:toy-pca} and this time, it shows that \texttt{search\_better} finds the optimum quicker than \texttt{search\_geodesic}.

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-tour-1}

@@

\caption{A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.}\label{fig:toy-tour}
\end{figure}

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original \(p \times d\) dimension enables a more holistic stereoscopic view of the search. To view a high dimensional (\(d \ge 3\)) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure \ref{fig:toy-tour} gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure \ref{fig:toy-tour} is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure \ref{fig:toy-pca}.

@@

In section \ref{toy-interp}, an interpolation with increase-then-decrease pattern has been presented. This pattern is undesirable since the optimiser could have started the next iteration from the highest basis on the tour path, as annotated as the interpolated basis in the plot, but instead, it is forced to start from the target basis. This motivates the design of an interruption to check the index value on the tour path so that the interpolating bases is accepted only up to the one with the largest index value. After implementing this interruption, the search finds a higher final index value with fewer steps as shown in the right panel of Figure \ref{fig:interruption}.

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/interruption-1}

@@

\caption{Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_better}. A comparison of the trace plot with and without interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path.}\label{fig:interruption}
\end{figure}

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/polish-1}

\caption{Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish.}\label{fig:polish}
\end{figure}

@@

Figure \ref{fig:polish} presents the final projections found before and after applying \texttt{search\_polish} on \texttt{search\_geodesic}. Polish search improves the index value from 0.9618 to 0.9627 with reduction of weights on the non-informative variables. In terms of the projected data as in Figure \ref{fig:polish}, polish works to sharpen the edges of each cluster.

@@

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/noisy-better-geo-1}

\caption{One-D projection on \texttt{boa5} data with noisy index \texttt{norm\_kol} optimised by \texttt{search\_geodesic} and \texttt{search\_better}. The grey dashed line represents the index value of the theoretical best basis. \texttt{search\_geodesic} fails to optimise the noisy index while \texttt{search\_better} has made reasonable improvements to reach the index value close to the one of the theoretical best basis.}\label{fig:noisy-better-geo}
\end{figure}

@@

Figure \ref{fig:noisy-better-geo} compares the tracing plot of two optimisers: \texttt{search\_geodesic} and \texttt{search\_better}. This time, the interpolated path is no longer smooth when using either algorithm and \texttt{search\_geodesic} fails to optimise this index with barely any improvement of the index value. On the other hand, \texttt{search\_better} is doing relatively well. With the theoretical best basis {[}0, 1, 0, 0, 0{]} producing an index value of 0.182, \texttt{search\_better} finds the final basis {[} 0.0376, -0.9916, -0.0581, -0.0831, 0.0716{]} with an index value of 0.165. A further polish step will give a marginal improvement of index value to 0.175 with a basis of {[} 0.0223, -0.9965, -0.0352, -0.0591, 0.0418{]}. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-better-1}

@@

\caption{One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better} with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced to 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points represent the local maximum corresponds to basis when \texttt{V2} is found and the global maximum where \texttt{V7} is found. The points produced by the search algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point.}\label{fig:kol-better}
\end{figure}

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-random-1}

@@

\caption{One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better\_random} with the same seeds. \texttt{search\_better\_random} has a probabilistic acceptance implementation that would also accept a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more instances find the global maximum.}\label{fig:kol-random}
\end{figure}

\begin{figure}

{\centering \includegraphics{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-random-tuned-1}

@@

\caption{One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better\_random} with an larger searching neighbourhood of 0.7. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter \texttt{alpha} from 0.5 to 0.7.}\label{fig:kol-random-tuned}
\end{figure}

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is \texttt{boa6} where \texttt{x2} and \texttt{x7} are informative. The two theoretical best bases are {[}0, 1, 0, 0, 0, 0{]} and {[}0, 0, 1, 0, 0, 0{]} with index value 0.182 and 0.235, respectively. Hence, the global maximum happens when variable \texttt{x7} is found. Simulation is done with 20 randomly generated seeds on two optimisers: \texttt{search\_better} (Figure \ref{fig:kol-better}) and \texttt{search\_better\_random} (Figure \ref{fig:kol-random} and Figure \ref{fig:kol-random-tuned}). Compared to Figure \ref{fig:kol-better} and \ref{fig:kol-random}, Figure \ref{fig:kol-random-tuned} further increases the neighbourhood parameter \texttt{alpha} from 0.5 to 0.7 to enlarge the search space. The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

In Figure \ref{fig:kol-better}, the starting point largely affects whether \texttt{search\_better} will find the global maximum. The three seeds on row three: seed 9145, seed 2511, and seed 9209, and three on row four: seed 2888, seed 9334, and seed 9819 easily find \texttt{V7} because they are born in Rome. On the other end, the ones on the first two rows can only find \texttt{V2}, largely due to the fact that they do not even have chances to search near \texttt{V7}.

A comparison between \texttt{search\_better} and \texttt{search\_better\_random} shows all four combinations where both, neither, or one of the two algorithms find the global maximum. Two iconic cases that shows how \texttt{search\_better\_random} improves \texttt{search\_better} are seeds 4761 (position (1,4) in Figure \ref{fig:kol-better} and (2,2) in Figure \ref{fig:kol-random}) and seed 1842 ((2,2) in Figure \ref{fig:kol-better} and (4,5) in Figure \ref{fig:kol-random}). In both cases, \texttt{search\_better\_random} accepts inferior points at the initial iteration and this acceptance has later been proved to change the direction of the search and hence allows the optimiser to explore near the global optimum and in the end, find the global optimum. However, this change of search direction may sometimes have an adverse effect on \texttt{search\_better\_random} and change the search to the neighbourhood of \texttt{V2}. This happens on seed 9209 and seed 9982 (position (1,1) and (1,2) in Figure \ref{fig:kol-random}). The two cases that neither algorithm finds the global optimum happens on seed 6170 and seed 2757 (position (1,1) and (1,2) in Figure \ref{fig:kol-better} and position (1,5) and (1,3) in Figure \ref{fig:kol-random}). This is largely due to the fact that the searching neighbourhood is not large enough to allow the space near \texttt{V7} to be adequately explored.

In Figure @ref\{fig:kol-random-tuned\}, the neighbourhood parameter \texttt{alpha} is increased from a 0.5 default to 0.7 to solve the insufficiency of search neighbourhood. Remember that a candidate basis is generated via a linear combination of the current basis and a randomly generated basis on the surface of the sphere, an increase of \texttt{alpha} gives more weights on the randomly generated bases and hence allows a wider search. Seed 9982 ((2,4) in Figure \ref{fig:kol-random-tuned}) and seed 2463 ((3,2) in Figure \ref{fig:kol-random-tuned}) benefit from this and find the global maximum. A noticeable feature of this enlarged search space is that the interpolation paths start to jump around the space but its usefulness is arguable. In seed 2888 (position (3,1)), seed 2986 (position (2,3)), seed 9209 (position (4,4)), just to name a few, this allows a near \texttt{V2} bases to switch to a near \texttt{V7} bases and as a result these cases find the global maximum. While in the case of seed 6746 (position (1,2)) and seed 2757 (position (1,5)), the jump happens from near \texttt{V7} basis to near \texttt{V2} basis, it does not come back and leaves those simulations in the local maximum.

@@

\begin{itemize}
\item
  \texttt{explore\_trace\_search()}: produces summary plots, as shown in Figure \ref{fig:toy-search}
\item
  \texttt{explore\_trace\_interp()}: produces trace plots for the interpolation points, as shown in Figure \ref{fig:toy-interp}
\item
  \texttt{explore\_space\_pca()}: produces plots of projection basis on the reduced space by PCA, as shown in Figure \ref{fig:toy-pca}. Animated version in Figure \ref{fig:toy-pca-animated} can be turned on via the argument \texttt{animate\ =\ TRUE}
\item
  \texttt{explore\_space\_tour()}: produces animated tour view on the full space of the projection bases, as shown in Figure \ref{fig:toy-tour}.
\item
  \texttt{get\_*()} extracts and manipulates certain components from the existing data object.

@@

\bibliographystyle{agsm}
\bibliography{biblio.bib}

\end{document}

>>>>

% !TeX root = RJwrapper.tex
\title{Visual diagnostics for constrained optimisation with application to guided tours}
\author{by H.Sherry Zhang}

\maketitle

\abstract{%
Projection pursuit searches for interesting low-dimensional views of high-dimensional data via the optimisation of an index function. The initial paper by Friedman \& Tukey in 1974 stated that ``the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure.'' However, while many projection pursuit indices have been proposed in the literature, less work has been done on the optimisation procedures. In this paper we introduce visual diagnostics for optimisation algorithms, in particular those available in the projection pursuit guided tour. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, ferrn, has been created to implement the diagnostics.
}

@@

Mathematical details of the geodesic interpolation can be found in \citet{buja2005computational}. Figure @ref(fig:tour-path) shows a sketch of the tour path. The blue frames are produced by the projection pursuit optimisation algorithm, and the white frames interpolate between them. The tour method has been implemented in the \texttt{tourr} package in R, available on the Comprehensive R Archive Network at \url{https://cran.r-project.org/web/packages/tourr/} \citep{wickham2011tourrpackage}.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=0.6\linewidth,height=0.3\textheight]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/img/tour_path}

@@

\caption[(ref:tour-path)]{(ref:tour-path)}(\#fig:tour-path)
\end{figure}
\end{Schunk}

(ref:tour-path) Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

\caption{\label{fig:toy-search} A comparison of search\_better with different values of the parameter max\_tries. A six-variable dataset boa6 is used with holes index on a 2D problem. max\_tries = 500 allows the optimiser to find better basis with higher index value at iteration six.}(\#fig:toy-search)
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

\caption[(ref:toy-interp)]{(ref:toy-interp)}(\#fig:toy-interp)
\end{figure}
\end{Schunk}

(ref:toy-interp) The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies.

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-1}

@@

\caption[(ref:toy-pca)]{(ref:toy-pca)}(\#fig:toy-pca)
\end{figure}
\end{Schunk}

(ref:toy-pca) 1D projection on the 5-variable dataset \texttt{boa5} with two optimisers: \texttt{search\_better} and \texttt{search\_geodesic}. The yellow point corresponds to the theoretical best basis {[}0, 1, 0, 0, 0{]} with \texttt{V2} being the only non-normal variable in the dataset. The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \texttt{search\_geodesic} has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.

@@

Figure @ref(fig:toy-pca) plots the first two principal components of two search paths, one using \texttt{search\_better} and the other using \texttt{search\_geodesic}. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in \texttt{search\_better} is broader, controlled by \texttt{alpha} parameter, which is default to 0.5 while the directional search in \texttt{search\_geodesic}, controlled by \texttt{delta} with a default of 0.01, is so tiny that it can barely be seen.

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

\caption[(ref:toy-pca-animated)]{(ref:toy-pca-animated)}(\#fig:toy-pca-animated)
\end{figure}
\end{Schunk}

(ref:toy-pca-animated) A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm.

Animated plots can be informative in diagnostics, especially in the case of PCA plot when the starting and ending of the search is not clear. Figure @ref(fig:toy-pca-animated) shows six frames of an animated version of Figure @ref(fig:toy-pca) and this time, it shows that \texttt{search\_better} finds the optimum quicker than \texttt{search\_geodesic}.

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

\caption[(ref:toy-tour)]{(ref:toy-tour)}(\#fig:toy-tour)
\end{figure}
\end{Schunk}

(ref:toy-tour) A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original \(p \times d\) dimension enables a more holistic stereoscopic view of the search. To view a high dimensional (\(d \ge 3\)) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure @ref(fig:toy-tour) gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure @ref(fig:toy-tour) is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure @ref(fig:toy-pca).

@@

In section \ref{toy-interp}, an interpolation with increase-then-decrease pattern has been presented. This pattern is undesirable since the optimiser could have started the next iteration from the highest basis on the tour path, as annotated as the interpolated basis in the plot, but instead, it is forced to start from the target basis. This motivates the design of an interruption to check the index value on the tour path so that the interpolating bases is accepted only up to the one with the largest index value. After implementing this interruption, the search finds a higher final index value with fewer steps as shown in the right panel of Figure @ref(fig:interruption).

(ref:interruption) Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_better}. A comparison of the trace plot with and without interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

\caption[(ref:polish)]{(ref:polish)}(\#fig:polish)
\end{figure}
\end{Schunk}

(ref:polish) Two-D projection on \texttt{boa6} data with holes index optimised by \texttt{search\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish.

@@

Figure @ref(fig:polish) presents the final projections found before and after applying \texttt{search\_polish} on \texttt{search\_geodesic}. Polish search improves the index value from 0.9618 to 0.9627 with reduction of weights on the non-informative variables. In terms of the projected data as in Figure @ref(fig:polish), polish works to sharpen the edges of each cluster.

@@

(ref:noisy-better-geo) One-D projection on \texttt{boa5} data with noisy index \texttt{norm\_kol} optimised by \texttt{search\_geodesic} and \texttt{search\_better}. The grey dashed line represents the index value of the theoretical best basis. \texttt{search\_geodesic} fails to optimise the noisy index while \texttt{search\_better} has made reasonable improvements to reach the index value close to the one of the theoretical best basis.

@@

Figure @ref(fig:noisy-better-geo) compares the tracing plot of two optimisers: \texttt{search\_geodesic} and \texttt{search\_better}. This time, the interpolated path is no longer smooth when using either algorithm and \texttt{search\_geodesic} fails to optimise this index with barely any improvement of the index value. On the other hand, \texttt{search\_better} is doing relatively well. With the theoretical best basis {[}0, 1, 0, 0, 0{]} producing an index value of 0.17, \texttt{search\_better} finds the final basis {[} 0.0376, -0.9916, -0.0581, -0.0831, 0.0716{]} with an index value of 0.165. A further polish step will give a marginal improvement of index value to 0.175 with a basis of {[} 0.0223, -0.9965, -0.0352, -0.0591, 0.0418{]}. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-better-1}

@@

\caption[(ref:kol-better)]{(ref:kol-better)}(\#fig:kol-better)
\end{figure}
\end{Schunk}

(ref:kol-better) One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better} with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced to 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points represent the local maximum corresponds to basis when \texttt{V2} is found and the global maximum where \texttt{V7} is found. The points produced by the search algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-random-1}

@@

\caption[(ref:kol-random)]{(ref:kol-random)}(\#fig:kol-random)
\end{figure}
\end{Schunk}

(ref:kol-random) One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better\_random} with the same seeds. \texttt{search\_better\_random} has a probabilistic acceptance implementation that would also accept a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more instances find the global maximum.

\begin{Schunk}
\begin{figure}

{\centering \includegraphics{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-random-tuned-1}

@@

\caption[(ref:kol-random-tuned)]{(ref:kol-random-tuned)}(\#fig:kol-random-tuned)
\end{figure}
\end{Schunk}

(ref:kol-random-tuned) One-D projection of \texttt{norm\_kol} index on \texttt{boa6} data optimised by \texttt{search\_better\_random} with an larger searching neighbourhood of 0.7. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter \texttt{alpha} from 0.5 to 0.7.

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is \texttt{boa6} where \texttt{x2} and \texttt{x7} are informative. The two theoretical best bases are {[}0, 1, 0, 0, 0, 0{]} and {[}0, 0, 1, 0, 0, 0{]} with index value 0.176 and 0.235, respectively. Hence, the global maximum happens when variable \texttt{x7} is found. Simulation is done with 20 randomly generated seeds on two optimisers: \texttt{search\_better} (Figure @ref(fig:kol-better)) and \texttt{search\_better\_random} (Figure @ref(fig:kol-random) and Figure @ref(fig:kol-random-tuned)). Compared to Figure @ref(fig:kol-better) and @ref(fig:kol-random), Figure @ref(fig:kol-random-tuned) further increases the neighbourhood parameter \texttt{alpha} from 0.5 to 0.7 to enlarge the search space. The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

In Figure @ref(fig:kol-better), the starting point largely affects whether \texttt{search\_better} will find the global maximum. The three seeds on row three: seed 9145, seed 2511, and seed 9209, and three on row four: seed 2888, seed 9334, and seed 9819 easily find \texttt{V7} because they are born in Rome. On the other end, the ones on the first two rows can only find \texttt{V2}, largely due to the fact that they do not even have chances to search near \texttt{V7}.

A comparison between \texttt{search\_better} and \texttt{search\_better\_random} shows all four combinations where both, neither, or one of the two algorithms find the global maximum. Two iconic cases that shows how \texttt{search\_better\_random} improves \texttt{search\_better} are seeds 4761 (position (1,4) in Figure @ref(fig:kol-better) and (2,2) in Figure @ref(fig:kol-random)) and seed 1842 ((2,2) in Figure @ref(fig:kol-better) and (4,5) in Figure @ref(fig:kol-random)). In both cases, \texttt{search\_better\_random} accepts inferior points at the initial iteration and this acceptance has later been proved to change the direction of the search and hence allows the optimiser to explore near the global optimum and in the end, find the global optimum. However, this change of search direction may sometimes have an adverse effect on \texttt{search\_better\_random} and change the search to the neighbourhood of \texttt{V2}. This happens on seed 9209 and seed 9982 (position (1,1) and (1,2) in Figure @ref(fig:kol-random)). The two cases that neither algorithm finds the global optimum happens on seed 6170 and seed 2757 (position (1,1) and (1,2) in Figure @ref(fig:kol-better) and position (1,5) and (1,3) in Figure @ref(fig:kol-random)). This is largely due to the fact that the searching neighbourhood is not large enough to allow the space near \texttt{V7} to be adequately explored.

In Figure @ref\{fig:kol-random-tuned\}, the neighbourhood parameter \texttt{alpha} is increased from a 0.5 default to 0.7 to solve the insufficiency of search neighbourhood. Remember that a candidate basis is generated via a linear combination of the current basis and a randomly generated basis on the surface of the sphere, an increase of \texttt{alpha} gives more weights on the randomly generated bases and hence allows a wider search. Seed 9982 ((2,4) in Figure @ref(fig:kol-random-tuned)) and seed 2463 ((3,2) in Figure @ref(fig:kol-random-tuned)) benefit from this and find the global maximum. A noticeable feature of this enlarged search space is that the interpolation paths start to jump around the space but its usefulness is arguable. In seed 2888 (position (3,1)), seed 2986 (position (2,3)), seed 9209 (position (4,4)), just to name a few, this allows a near \texttt{V2} bases to switch to a near \texttt{V7} bases and as a result these cases find the global maximum. While in the case of seed 6746 (position (1,2)) and seed 2757 (position (1,5)), the jump happens from near \texttt{V7} basis to near \texttt{V2} basis, it does not come back and leaves those simulations in the local maximum.

@@

\begin{itemize}
\item
  \texttt{explore\_trace\_search()}: produces summary plots, as shown in Figure @ref(fig:toy-search)
\item
  \texttt{explore\_trace\_interp()}: produces trace plots for the interpolation points, as shown in Figure @ref(fig:toy-interp)
\item
  \texttt{explore\_space\_pca()}: produces plots of projection basis on the reduced space by PCA, as shown in Figure @ref(fig:toy-pca). Animated version in Figure @ref(fig:toy-pca-animated) can be turned on via the argument \texttt{animate\ =\ TRUE}
\item
  \texttt{explore\_space\_tour()}: produces animated tour view on the full space of the projection bases, as shown in Figure @ref(fig:toy-tour).
\item
  \texttt{get\_*()} extracts and manipulates certain components from the existing data object.

@@

\bibliography{RJreferences}

\address{%
H.Sherry Zhang\\
Department of Econometrics and Business Statistics, Monash University\\
\\
}
\href{mailto:Huize.Zhang@monash.edu}{\nolinkurl{Huize.Zhang@monash.edu}}
